Data Specialization for a Staged Lambda Calculus

### 2014-12-01 paper/implementation.tex
added a more about mono

2`l => 
letfun double (n : unit, l0 : unit) = ()
in (double (pi2 l); pi1 l)-100`
\end{lstlisting}

@@

We choose to investigate the latter approach.
Specifically, we add new staging features to \lang\ which allow us to indicate that a piece of code is {\em strictly} 
stage \bbone---{\em i.e.} that it contains no $\next$ blocks.
At the type level, this manifests as a constructor called $\monoType$.
So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
which might fail to terminate at stage \bbtwo.  
Alternatively $\monoType (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
which {\em cannot} have any content at stage \bbtwo.
Naturally, the well-formedness of $\monoType A$ requires that $A$ contain no $\fut$s.

At the term level, we add the $\monoTerm$, $\lifttag$, \texttt{let mono} constructs to move information
in and out of $\monoType$.

>>>>

2`l => 
letfun double (n : unit, l0 : unit) = ()
in (double (pi2 l); pi1 l)-7`
\end{lstlisting}

@@

We choose to investigate the latter approach.
Specifically, we add new staging features to \lang\ which allow us to indicate that a piece of code is {\em strictly} 
stage \bbone---{\em i.e.} that it contains no $\next$ blocks.
At the type level, this manifests as a constructor called $\monoType$.
So for instance, $\rmint \to \rmint$ is a stage \bbone\ function from one integer to another,
which might fail to terminate at stage \bbtwo,  
but $\monoType (\rmint \to \rmint)$ is a stage \bbone\ function from one integer to another
which {\em cannot} have any content at stage \bbtwo.
Naturally, the well-formedness of $\monoType A$ requires that $A$ contain no $\fut$s.

At the term level, we add the 
$\monoTerm$, \texttt{letmono}, $\lifttag$, and $\liftint$ 
constructs to move information in and out of $\monoType$.  
The statics, dynamics, and splitting rules for these constructs are given in \cref{fig:monoRules}.

The $\monoTerm$ construct allows us to declare a block of code as entirely stage \bbone,
equivalent to saying that it has no $\next$ blocks.
This is enforced in the type system by checking at the new world $\bbmono$ within the $\monoTerm$ block.
Accordingly, $\monoTerm$ blocks are the intro form for the $\monoType$ modality.

The \texttt{letmono} construct allows us to form let bindings of monostage code without going into a $\monoTerm$ block.
This construct, along with $\monoTerm$ blocks, is enough to allow us to form terms of the following types:
\begin{itemize}
\item $\monoType(A \times B) \to \monoType A \times \monoType B$ 
\item $\monoType A \times \monoType B \to \monoType(A \times B)$
\item $\monoType A + \monoType B \to \monoType(A + B)$
\item $\monoType(A \to B) \to (\monoType A \to \monoType B)$
\end{itemize}
There are two notable types absent in this list, namely the converses of the latter two entries.
Lets double-check to make sure that this is the behavior we'd want.

First consider $(\monoType A \to \monoType B) \to \monoType(A \to B)$.
Were this type inhabited, it would suggest that any function which might not terminate at stage $\bbtwo$
can be transmuted into a function that does all the same work but definately terminates at stage $\bbtwo$.
That seems definitely wrong, so it makes sense that this type is uninhabited.

Next, consider $\monoType(A + B) \to \monoType A + \monoType B$.
That would suggest that if there's a sum tag computed in a monostage way,
then that sum tag can be lifted out from under the $\monoType$.
This feature {\em does} sound reasonable, and so we add it explicitly
with the $\lifttag$ construct. 
Similarly, we add $\liftint$ to move integers out from under $\monoType$.
\footnote{As with $\pause$, $\liftint$ can be defined in the language with
some effort, but we provide it as a convenience.}

With these new features in place, we can now define the \texttt{double}
function so that it splits reasonably.
\begin{lstlisting}
2`prev{
  1`let mono{`3`double`1`} = mono{`3`fn n : int => n+n`1`}
  in next{`2`hold {`1`lift(mono{`3`double 5`1`})`2`}-7`1`}`2`
}`
\end{lstlisting}
This now splits into:
\begin{lstlisting}
1`let double = (fn n : int => n+n)
in (double 5,())`

2`l => in (pi1 l)-7`
\end{lstlisting}

### 2014-12-01 paper/paper.tex
uncomment the splitting section
worked on global structure of splitting section
minor tweaks
small changes; lots of commenting-out
rewrite 4.0 - 4.1 with nico
comments entered
minor
migrate overview section's text into semantics section
some work on the mono section
added a more about mono

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{bbm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
%\usepackage{mathabx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{cprotect}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
%\usepackage[utf8x]{inputenc}
%\PrerenderUnicode{λ}
\usepackage{verbatim}

\usepackage{cleveref}
\AtBeginDocument{\renewcommand\ref\cref}
\crefname{figure}{Figure}{Figures}
\crefname{theorem}{Theorem}{Theorems}

@@

\definecolor{stageOneColor}{RGB}{120,30,0}
\definecolor{stageTwoColor}{RGB}{0,0,127}

@@

\lstset{%language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
%        escapeinside={"*}{*"},
        escapeinside={@}{@},
        moredelim=*[is][\color{stageOneColor}]{1`}{`},
        moredelim=*[is][\color{stageTwoColor}]{2`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

@@

\title{Stage-Splitting a Modal Language}

@@

@@

%% %\vfill\eject\pagebreak
%%\input{splitting}
\input{algoexamples}
\input{relatedwork}
\input{umut}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{proof}
\usepackage{mathpartir}
\usepackage{bbm}
%\usepackage{mathabx}
\usepackage{listings}
\usepackage{enumitem}
\usepackage{url}
\usepackage[colorlinks=true,allcolors=blue,breaklinks,draft=false,unicode=true]{hyperref}
\usepackage[usenames,dvipsnames]{color}
\usepackage{cprotect}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
%\usepackage[utf8x]{inputenc}
%\PrerenderUnicode{λ}
\usepackage{verbatim}

\usepackage{cleveref}
\AtBeginDocument{\renewcommand\ref\cref}
\crefname{section}{Section}{Sections}
\crefname{figure}{Figure}{Figures}
\crefname{theorem}{Theorem}{Theorems}

@@

\definecolor{stageOneColor}{RGB}{120,30,0}
\definecolor{stageTwoColor}{RGB}{0,0,127}
\definecolor{stageMonoColor}{RGB}{0,120,0}

@@

\lstset{%language=ML,
        columns=fullflexible,
        keepspaces=true,
        basicstyle=\ttfamily,
        tabsize=4,
%        escapeinside={"*}{*"},
        escapeinside={@}{@},
        moredelim=*[is][\color{stageOneColor}]{1`}{`},
        moredelim=*[is][\color{stageTwoColor}]{2`}{`},
        moredelim=*[is][\color{stageMonoColor}]{3`}{`},
        literate={|->}{{\ $\mapsto$\ }}1
                 {xhat}{{$\hat{\texttt{x}}$}}1
                 {yhat}{{$\hat{\texttt{y}}$}}1
                 {zhat}{{$\hat{\texttt{z}}$}}1
                 {murec}{{$\mu$}}1
                 {alpha}{{$\alpha$}}1
                 {beta}{{$\beta$}}1
                 {$}{{$\fut$}}1}

@@

\title{Data Specialization for a Staged Lambda Calculus}

@@

\section{Intro}
(placeholder in order to keep section numbers stable)

In \ref{fig:termSplitSummary}, we summarize the terminology of our
formal development.

@@

%% %\vfill\eject\pagebreak
\input{splitting}
\input{implementation}
\input{algoexamples}
\input{relatedwork}
\input{umut}
